---
title: "Introduction to Spatial Data Types in R"
author: "claudia a engel"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_notebook:
    code_folding: hide
    toc: yes
    toc_depth: 4
---
```{r setup, include=FALSE}
## Global code options
knitr::opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)

## libraries needed for R code examples
library(sp)
library(raster)
```

***

For this module you need to have the [`sp`](https://CRAN.R-project.org/package=sp) and [`raster`](https://CRAN.R-project.org/package=raster) libraries installed and loaded. [Note that the latest `raster`(2.5-8) depends on `sp` 1.2-0 or later, so make sure that your versions align.]

__Mac users__ 

Since there is now a Mac binary package for `rgdal` [available on CRAN](https://CRAN.R-project.org/package=rgdal) installing `sp` with the default settings will also install that `rgdal` as one of its dependencies. It includes a built-in basic GDAL that lacks some of the extra GDAL formats can handle. There is an alternative `rgdal` package distributed by kyngchaos that you can use instead. Formats **not** included in the CRAN distribution but included in the kyngchaos distribution are currently: DODS, Geomedia, Interlis 1, Interlis 2, LIBKML, MSSQLSpatial, NAS, ODBC, OGDI, PGeo, PostgreSQL, OSI, Walk, XLS. (You mostly might care about having the PostgreSQL driver at some point.)


If you __DON'T__ care, simply say:

    install.packages(c("sp", "raster"))

If you __DO__ care, follow the instructions below.

#. Install the raster library: 

        install.packages("raster"))

#. Don't install `sp` with dependencies: 

        install.packages("sp", dependencies = F)

#. Download the latest GDAL complete (GDAL 2.1) from [this site](http://www.kyngchaos.com/software/frameworks)

#. Doubleclick and install the downloaded `.dmg` file as you are used to on a Mac.

#. Make sure you have R Version 3.3 or later installed -- if not update it.

#. Download a different rgdal (rgdal 1.1.10-1) from [the same site](http://www.kyngchaos.com/software/frameworks).  

#. Doubleclick to open the `.dmg` file

#. Move `rgdal_*.tgz` to your Desktop folder

#. Install the local package with:

        install.packages("~/Desktop/rgdal_*.tgz", repos = NULL, type = .Platform$pkgType)

__Windows users__

    install.packages(c("sp", "raster"))
    
__Mac and Windows__

Test if all went well:

    library(sp)
    library(raster)

***

## 1. Spatial objects in R

### Conceptualizing a spatial Object

In vector GIS we deal with, points, lines, and polygons, like so:

```{r echo=FALSE}
px <- c(5, 7, 8, 9, 8, 7, 6) 
py <- c(7, 3, 4, 8, 9, 15, 14) 
plot(px, py, type="n", axes=F, xlab = '', ylab = '') 
polygon(px, py)
points(c(6, 9, 8, 8.5), c(9, 14, 8, 9), pch=20)
lines(c(5, 6, 7, 8), c(5, 6,10, 11))
lines(c(8, 9), c(14, 12))
```

***

#### Exercise 1

Discuss with your neighbor: What parameters do we need to specify in order to describe spatial vector objects?

```{r}
# * lat/lon coordinates
# * projection
# * attribute data
# * if polygon, is it a hole or not
# * ... ?
```


***

In R the `sp` package provides classes and methods for spatial data types[^1].

[^1]: From R Bivand (2011) [Introduction to representing spatial objects in R](http://geostat-course.org/system/files/monday_slides.pdf)

Development of the `sp` package began in the early 2000s in an attempt to standardize how spatial data would be treated in R and to allow for better interoperability between different analysis packages that use spatial data. The package provides classes and methods to create _points_, _lines_, _polygons_, and _grids_ and to operate on them. It is one of the most important packages that you will need to use if dealing with spatial data in R. Many of the spatial analysis packages now use the spatial data types that are implemented in `sp` i.e. they "depend" on the `sp` package.

In `sp` spatial objects are conceptualized in the following way[^2]:

[^2]: Note that this is not the only way spatial objects are conceptualized in R. Other spatial packages may use their own class definitions for spatial data (for example `spatstat`). Usuallly you can find functions that convert `sp` objects to and from these formats.

The foundational structure for *any* spatial object in `sp` is the `Spatial` class. It has two slots (new-style class objects in R have pre-defined components called slots):

* a __bounding box__ 
      
* a __CRS class object__ to define the Coordinate Reference System 

## 2. Creating a spatial object

In order to create a spatial object manually the basic steps are:  

> I. Create a bunch of points, lines, or polygons (details below)
 
> II. Convert those to a `Spatial*` object (`*` stands for Points, Lines, or Polygons). This steps adds the bounding box (automatically) and the slot for the Coordinate Reference System or CRS (which needs to be filled with a value manually). 
 
> III. (_Optional_:) Add a data frame with attribute data, which will turn your `Spatial*` object into a `Spatial*DataFrame` object.

### I. Create geometric objects (topology)

__Points__ (which may have 2 or 3 dimensions) are the most basic spatial data objects. They are generated out of either a single coordinate or a set of coordinates[^3], like a two-column matrix or a dataframe with a column for latitude and one for longitude. 

[^3]: Coordinates should be of type double and will be promoted if not.

__Lines__ are generated out of `Line` objects. A `Line` object is a spaghetti collection of 2D coordinates and is generated out of a two-column matrix or a dataframe with a column for latitude and one for longitude. A `Lines` object is a __list__ of one or more `Line` objects, for example all the contours at a single elevation.

__Polygons__ are generated out of `Polygon` objects. A `Polygon` object is a spaghetti collection of 2D coordinates with equal first and last coordinates and is generated out of a two-column matrix or a dataframe with a column for latitude and one for longitude. A `Polygons` object is a __list__ of one or more `Polygon` objects, for example islands belonging to the same country.


### II. Create spatial objects
`SpatialPoints` can be directly generated out of the coordinates. Lines and polygons are part of the R's basic `graphics` package used for plotting, but we need to turn those into spatial, i.e. "geographically aware" objects. The `sp` package provides functions do do that. `SpatialLines` and `SpatialPolygons` objects are generated using lists of `Lines` or `Polygons` objects respectively (more below).

### III. Add attributes
The points in a `SpatialPoints` object may be associated with a row of attributes to create a `SpatialPointsDataFrame` object. The coordinates and attributes may, but do not have to be keyed to each other using ID values.

`SpatialLinesDataFrame` and `SpatialPolygonsDataFrame` objects are defined using `SpatialLines` and `SpatialPolygons` objects and data frames. The ID fields are here required to match the data frame row names.

| How | to | create | spatial | objects: |
| --- | --- | --- | --- | --- |
| matrix or data frame with coordinates | &#x2192; | &#x2192; | SpatialPoints(coordinates) | SpatialPointsDataFrame(SpatialPoints object and data frame) |
| matrix or data frame with coordinates | Line(coordinates) | Lines(list of Line objects) | SpatialLines(list of Lines objects) | SpatialLinesDataFrame(SpatialLines object and data frame, matched by ID) |
| matrix or data frame with coordinates | Polygon(coordinates) | Polygons(list of Polygon objects) | SpatialPolygons(list of Polygons objects) | SpatialPolygonsDataFrame(SpatialPolygons object and data frame, matched by ID) |


### Spatial methods

A number of spatial methods are available for the classes in `sp`. Among the ones I often use are:

function | and what it does
------------ | ------------------------------------------------------
`bbox()` | returns the bounding box coordinates
`proj4string()` | sets or retrieves projection attributes using the CRS object.
`CRS()` | creates an object of class of coordinate reference system arguments
`spplot()` | plots a separate map of all the attributes unless specified otherwise
`coordinates()` | set or retrieve the spatial coordinates. For spatial polygons it returns the centroids.
`over(a, b)` | used for example to retrieve the polygon or grid indexes on a set of points - we'll come back to that one later
`spsample()` | sampling of spatial points within the spatial extent of objects

### Creating a spatial object from scratch

***
#### Exercise 2

In this example we will manually generate a point vector object and plot it.

```{r}
xy <- matrix(runif(20), ncol=2) # a matrix with some arbitrary points as coordinates..
xy_sp <- SpatialPoints(xy) # ..converted into a spatial object
plot(xy_sp, pch = 19)
```

Test out some commands:
```{r}
coordinates(xy_sp)
```
```{r}
bbox(xy_sp)
```
```{r}
summary(xy_sp)
```
```{r}
str(xy_sp)
```

Add attributes:
```{r}
df <- data.frame(attr1 = c(LETTERS[1:5], LETTERS[1:5]), attr2 = rep(1:5, each = 2))
xy_spdf <- SpatialPointsDataFrame(xy_sp, df)
summary(xy_spdf)
```

Some subsetting:
```{r}
xy_spdf$attr2    # column with "attr2" only -- this is a vecor
```

```{r}
xy_spdf[xy_spdf$attr1 == "A", ]    # selection of attr1 -- this is an SP object
```

***
### Creating a spatial object from a lat/lon table

A `SpatialPointsDataFrame` object can be created directly from a table by specifying which columns contain the coordinates. For example, you might have a spreadsheet that contains latitude, longitude and some attribute values. You can read the spreadsheet into a data frame with `read.table` or `read.csv` and create the SP object from the data frame in one step by using the `coordinates()` function. As mentioned above that function can be used to retrieve spatial coordinates but also to set them, which is done in R fashion with:

    coordinates(myDataframe) <- value

`value` can have different forms -- in this example needs to be a character vector which specifies the data frame's columns for the longitude and latitude (x,y) coordinates.

If we use this on a data frame it automatically converts the data frame object into a `SpatialPointsDataFrame` object.

***
####Exercise 3

1. Download and unzip [`RSpatialDataTypes.zip`](https://www.dropbox.com/s/g5p8b1xi2k5lydw/RSpatialDataTypes.zip?dl=1)
2. Use `read.csv()` to read `PhiladelphiaZIPHousing.csv` into a dataframe in R and name it `ph_df`.
3. Use `head()` to examine the first few lines of the dataframe. What information does it contain?
4. Use `class()` to examine which object class the table belongs to.
5. Convert the table into a spatial object with using the `coordinates` function.
6. Use `class()`again to examine which object class the table belongs to now:  
What to you observe?
7. Plot, using the attributes from the table, for example:  
    `bubble(ph_df, "price")` or  
    `spplot(ph_df, "use")`

> Try before you peek! 

```{r eval=FALSE}
ph_df <- read.csv("~/Desktop/RSpatialDataTypes/PhiladelphiaZIPHousing.csv")
head(ph_df)
class(ph_df)
coordinates(ph_df) <- c("lon", "lat")
class(ph_df)
bubble(ph_df, "price")
```

### A brief, but important word about projection.

Note that the Spatial object you just created __does not__ have a projection defined. It is ok to plot, but be aware that for any meaningful spatial operation you will need to define a projection. 

9. This is how it's done:

```{r eval=FALSE}
is.projected(ph_df) # see if a projection is defined  
proj4string(ph_df) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs") # this is WGS84
is.projected(ph_df) # voila
```

***


The following example[^4] shows how a set of polygons are built from scratch. Note that the coordinates of the _poly4_ polygon move in the opposite direction (anti-clockwise) than the other three (clockwise); _poly4_ is meant to represent a hole in the _poly3_ polygon. The default value for the hole colour is "transparent". `Polygons` objects have to be given **character ID values that are unique** when combined to a `SpatialPolygons` object.

[^4]:adapted from [Edzer Pebesma Roger S. Bivand Feb 2005: S Classes and Methods for Spatial Data: the sp Package](http://cran.r-project.org/web/packages/sp/vignettes/intro_sp.pdf)


```{r}
# make up some coordinates, note that first and last coordinate pairs are identical
coords1 <- cbind(c(2,4,4,1,2),c(2,3,5,4,2))
coords2 <- cbind(c(5,4,2,5),c(2,3,2,2))
coords3 <- cbind(c(4,4,5,10,4),c(5,3,2,5,5))
coords4 <- cbind(c(5,6,6,5,5),c(4,4,3,3,4))

# create individual polygon objects from coordinates
poly1 <-  Polygon(coords1)
poly2 <-  Polygon(coords2)
poly3 <-  Polygon(coords3)
poly4 <-  Polygon(coords4, hole = TRUE)

# create lists of polygon objects from polygon objects and unique ID
polys1 <-  Polygons(list(poly1), "poly1")
polys2 <-  Polygons(list(poly2), "poly2")
polys3 <-  Polygons(list(poly3, poly4), "poly3/4")

# create spatial polygons object from lists
spatial_poly <-  SpatialPolygons(list(polys1, polys2, polys3), pO = 1:3) # pO: plotting order
plot(spatial_poly, col="dark green")
```

In order to add attributes to the polygons either the `row.names` of the attributes data frame need to be set to the Polygon ID values of the SpatialPolygons object or one column of the data frame needs to contain the matching Polygons IDs. The rows of the data frame will be re-ordered if necessary.

```{r}
attr1 <- data.frame(characters=LETTERS[1:3], row.names=c("poly1", "poly2", "poly3/4"))  # attribute values data frame with poly IDs as rownames
spatial_polyDf <- SpatialPolygonsDataFrame(spatial_poly, attr1) # combine

# this also works:
attr2 <- data.frame(characters=LETTERS[1:3], polyIDs=c("poly1", "poly2", "poly3/4"))
spatial_polyDf2 <- SpatialPolygonsDataFrame(spatial_poly, attr2, match.ID = "polyIDs") 

spplot(spatial_polyDf) 
# spplot(spatial_polyDf2, "characters")
```

Let's look at its structure:
```{r}
str(spatial_polyDf)
```
Whoa. 

Note that we can access the information in the slots using the `@`. For example: 
```{r}
spatial_polyDf@bbox
```

However, it is **strongly encouraged** to use the provided functions and methods instead, like:
```{r}
bbox(spatial_polyDf)
```

To look at the attribute table:
```{r}
as.data.frame(spatial_polyDf) # coerce into data frame instad of: spatial_polyDf@data
```

## 3. Getting spatial data in and out of R 

The good news is that usually we do not have to create `SpatialPolygonDataframe` objects as tediously as we did above. It is much more common that we work with already existing spatial data. 

### How to work with `rgdal`

In order to read spatial data into R and turn them into `Spatial*` family objects we heavily rely on the `rgdal` package. It provides us direct access to the powerful [GDAL library](http://gdal.org) from within R. 

We can read in and write out spatial data using:

    `readOGR()` and `writeOGR()` (for vector)  
    `readGDAL()` and `writeGDAL()` (for raster/grids)

The parameters provided for each function vary depending on the exact spatial file type you are reading. We will take an ESRI shapefile as an example. A shapefile - as you know - [consists of various files of the same name, but with different extensions](https://en.wikipedia.org/wiki/Shapefile). They should all be in one directory and that is what R expects.

When reading in a shapefile, `readOGR()` requires the following two arguments:

    datasource name (dsn)  # the path to the folder that contains the files
                           # this is a path to the folder, not a filename!
    layer name (layer)     # the shapefile name WITHOUT extension
                           # this is not a path but just the name of the file!

Setting these arguments correctly can be cause of much headache for beginners, so let me spell it out:

- Firstly, you obviously need to know the name of shapefile.

- Secondly, you need to know the name and location of the folder that contains all the shapefile parts.

- Lastly, `readOGR` only reads the file and spits it on your screen. But similarly to reading in csv tables you want to actually work with the file, so you need to assign it to an R object.

For example:

- I have a shapefile called `myShapefile.shp` and all its associated files (like _.dbf, .prj, .shx_, ...) in a directory called `myShapefileDir` in my desktop folder, 
- I have my R working directory set to my desktop folder,
- I want to assign the shape file to an R object called `myShape`.

Then my command to read this shapefile would look like this:
```
myShape <- readOGR(dsn = "myShapefileDir", layer = "myShapefile")
```
or in short:
```
myShape <- readOGR("myShapefileDir", "myShapefile")
```

Now let's do this.

***
####Exercise 4

1. Load the `rgdal` package.
2. Determine the location of the folder enclosing the `PhillyTotalPopHHinc` shapefile.
3. Read `PhillyTotalPopHHinc` into an object called `philly`. Make sure you provide the appropriate directory structure.
4. Examine the object, for example with `summary()` or `class()`
5. Plot it.
6. Take a look at the column names of the attribute data with `names()`
7. Take a look at the attribute data with `head()`
8. Select a subset of polygons with a median household income (`medHHinc`) of over 60000.  
9. Add that to the plot. In red.

> Try before you peek!

```{r eval=F}
library(rgdal)
philly <- readOGR("/Users/cengel/Desktop/RSpatialDataTypes/Philly/", "PhillyTotalPopHHinc") 
# side note: unlike read.csv readOGR does not understand the ~ as valid element of a path. This (on Mac) will not work:
# philly <- readOGR("~/Desktop/RSpatialDataTypes/Philly/", "PhillyTotalPopHHinc")
summary(philly)
class(philly)
names(philly)
head(philly)
plot(philly)
philly_rich <- philly[which(philly$medHHinc > 60000),]
plot(philly); plot(philly_rich, add=T, col="red")
```

***

GDAL supports over 200 [raster formats](http://www.gdal.org/formats_list.html) and [vector formats](http://www.gdal.org/ogr_formats.html). Use `ogrDrivers()` and `gdalDrivers()` (without arguments) to find out which formats your `rgdal` install can handle.


### Raster data

Raster files, as you probably know, have a much more compact data structure than vectors. Because of their regular structure the coordinates do not need to be recorded for each pixel or cell in the rectangular extent. A raster is defined by:

- a CRS 
- coordinates of its origin 
- a distance or cell size in each direction 
- a dimension or numbers of cells in each direction
- an array of cell values 

Given this structure, coordinates for any cell can be computed and don't need to be stored. 

In `sp` the `GridTopology` class is the key element of raster representations[^5]. It contains 

* the center coordinate pair of the south-west raster cell, 
* the two cell sizes in the metric of the coordinates, giving the step to successive centres, and 
* the numbers of cells for each dimension. 

[^5]: There is also a `SpatialPixels` object which stores grid topology and coordinates of the actual points.

A simple grid can be built like this:

```{r tidy=F}
# specify the grid topology with the following parameters:
# - the smallest coordinates for each dimension, here: 0,0
# - cell size in each dimension, here: 1,1 
# - number of cells in each dimension, here: 5,5
gtopo <- GridTopology(c(0,0), c(1,1), c(5,5)) # create the grid
datafr <- data.frame(runif(25)) # make up some data
SpGdf <- SpatialGridDataFrame(gtopo, datafr) # create the grid data frame
summary(SpGdf)
```

A very good alternative is the `raster` package, which works slightly differently.  
The `raster` package is a major extension of spatial data classes to access large rasters and in particular to process very large files. It includes object classes for `RasterLayer`, `RasterStacks`, and `RasterBricks`, functions for converting among these classes, and operators for computations on the raster data. Conversion from `sp` type objects into `raster` type objects is easy.

If we wanted to do the same as above, namely creating the same raster object from scratch we would do the following:

```{r tidy=F}
# specify the RasterLayer with the following parameters:
# - minimum x coordinate (left border)
# - minimum y coordinate (bottom border)
# - maximum x coordinate (right border)
# - maximum y coordinate (top border)
# - resolution (cell size) in each dimension
r <- raster(xmn=-0.5, ymn=-0.5, xmx=4.5, ymx=4.5, resolution=c(1,1))
r
```

So here we have created an object of type `RasterLayer`, as compared to above, where we created an object of type `GridTopology`.

Compare this to the output from above and __note something important here__: Different from the grid object we generated from scratch, this raster object has a CRS defined! If the crs argument is missing when creating the Raster object, the x coordinates are within -360 and 360 and the y coordinates are within -90 and 90, the WGS84 projection is used by default! 

Good to know.

To add some values to the cells we could the following. Be aware that different from the `GridTopology` object above, which we converted to a `SpatialGridDataFrame` when adding values, this object here remains a `RasterLayer`.

```{r tidy=F}
class(r)
r <- setValues(r, runif(25))
class(r)
plot(r); points(coordinates(r), pch=3)
```

(See the [`rasterVis` package](https://cran.r-project.org/web/packages/rasterVis/index.html) for more advanced plotting of `Raster*` objects.)

RasterLayer objects can also be created from a matrix.

```{r}
class(volcano)
volcano.r <- raster(volcano)
class(volcano.r)
```

To read in a raster file we can use `readGDAL()` from the `sp` package, which requires only the filename of the raster as argument.

The respective function in `raster` package is called `raster()`.

***

#### Exercise 5

1. Load the `raster` library
2. Read in the DEM using the `raster()` function
3. Examine by typing the name you gave the DEM 
4. Extract contour lines and plot them with `contour()`

> Try before you peek!

```{r eval=F}
library(raster)
dem.r <- raster("~/Desktop/RSpatialDataTypes/DEM_10m/bushkill_pa.dem")
dem.r
contour(dem.r)
```

***


[^6]: From (https://www.e-education.psu.edu/geog482fall2/c7_p8.html)

There are currently about 170 [R packages on CRAN for reading, visualising, and analysing (geographical) spatial data](http://cran.r-project.org/web/views/Spatial.html). I recommend to visit that website if you are exploring spatial analysis with R.